---
title: "Class 12 HW"
author: "May Alsoufi (PID: A17423250)"
format: pdf
---

## Section 4: Population Scale Analysis [HOMEWORK]

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. 

```{r}
dat <- read.table("rs8067378_ENSG00000172057.6.txt",
                  header = TRUE)

head(dat)
```

```{r}
str(dat)
```

```{r}
table(dat$geno)
```
```{r}
tapply(dat$exp, dat$geno, median)
```
Can also be shown like this:

```{r}
aggregate(exp ~ geno, dat, median)

```

The dataset contained 108 individuals with genotype A/A, 233 with A/G, and 121 with G/G.
The median expression levels were 31.2 for A/A, 25.1 for A/G, and 20.1 for G/G.
Expression decreases progressively with increasing copies of the G allele.

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
boxplot(exp ~ geno,
        data = dat,
        col = c("pink", "lightblue", "lavender"),
        main = "ORMDL3 Expression by rs8067378 Genotype",
        xlab = "Genotype",
        ylab = "Expression Level")
```

The boxplot shows  differences in ORMDL3 expression across genotypes. Individuals with the A/A genotype exhibit the highest median expression, followed by A/G, while G/G individuals show the lowest expression levels.

This pattern suggests an effect of the G allele where increasing copies of the G allele correspond to progressively lower ORMDL3 expression.

rs8067378 appears to influence ORMDL3 gene expression, indicating that this SNP likely functions as a regulatory variant.