---
title: "Class 18"
author: "May Alsoufi (PID: A17423250)"
format: pdf
toc: TRUE
---

### Background
Pertussis (more commonly known as whooping cough) is a highly contagious respiratory disease caused by the bacterium Bordetella pertussis.  People of all ages can be infected leading to violent coughing fits followed by a characteristic high-pitched “whoop” like intake of breath. Children have the highest risk for severe complications and death. Recent estimates from the WHO indicate that ~16 million cases and 200,000 infant deaths are due to pertussis annually.

### 1. Investigating pertussis cases by year
The United States Centers for Disease Control and Prevention (CDC) has been compiling reported pertussis case numbers since 1922 in their National Notifiable Diseases Surveillance System (NNDSS). We can view this data on the CDC website here: https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html

```{r}
library(tidyverse)
library(jsonlite)
library(lubridate)
library(datapasta)
```

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
library(rvest)

url <- "https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html"

tables <- url |>
  read_html() |>
  html_table()

cdc <- tables[[1]]
```

```{r}
cdc$cases <- as.numeric(gsub(",", "", cdc$`No. Reported Pertussis Cases`))
cdc$year <- cdc$Year

cdc <- cdc |> select(year, cases)
```

```{r}
ggplot(cdc, aes(year, cases)) +
  geom_point() +
  geom_line() +
  labs(title="Pertussis Cases by Year",
       x="Year",
       y="Number of cases")
```

### 2. A tale of two vaccines (wP & aP)
Two types of pertussis vaccines have been developed: whole-cell pertussis (wP) and acellular pertussis (aP). The first vaccines were composed of ‘whole cell’ (wP) inactivated bacteria. The latter aP vaccines use purified antigens of the bacteria (the most important pertussis components for our immune system). These aP vaccines were developed to have less side effects than the older wP vaccines and are now the only form administered in the United States.

> Q2. Using the ggplot `geom_vline()` function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine. What do you notice?

```{r}
ggplot(cdc, aes(x = year, y = cases)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = 1946, color = "blue") +
  geom_vline(xintercept = 1996, color = "red") +
  labs(title = "Pertussis Cases by Year",
       x = "Year",
       y = "Number of Cases")
```

After the introduction of the wP vaccine in 1946, the number of pertussis cases drops dramatically over the following few decades. By the 70s, reported cases are extremely low compared to earlier years. After the switch to the aP vaccine in 1996, the number of cases begins to increase again gradually over time. This is suggesting that while vaccination significantly reduced disease incidence, the newer aP vaccine may provide shorter-lasting immunity or different levels of protection compared to the earlier wP vaccine.

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the acellular pertussis (aP) vaccine in 1996, reported pertussis cases begin to increase again after a long period of low incidence. One possible explanation is that the aP vaccine produces weaker or shorter-lasting immunity than the earlier whole-cell (wP) vaccine. Because the acellular vaccine contains fewer bacterial components, immunity may wane more quickly, allowing susceptibility in vaccinated individuals and contributing to the observed rise in cases.

### 3. Exploring CMI-PB data
Why is this vaccine-preventable disease on the upswing? To answer this question we need to investigate the mechanisms underlying waning protection against pertussis. This requires evaluation of pertussis-specific immune responses over time in wP and aP vaccinated individuals.

## The CMI-PB API returns JSON data
The CMI-PB API (like most APIs) sends responses in JSON format. Briefly, JSON data is formatted as a series of key-value pairs, where a particular word (“key”) is associated with a particular value. To read these types of files into R we will use the read_json() function from the jsonlite package. Note that if you want to do more advanced querys of APIs directly from R you will likely want to explore the more full featured rjson package. The big advantage of using jsonlite for our current purposes is that it can simplify JSON key-value pair arrays into R data frames without much additional effort on our part.

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
subject <- read_json(
"https://www.cmi-pb.org/api/subject",
simplifyVector = TRUE)

head(subject)
```

```{r}
table(subject$infancy_vac)
```
> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex?

```{r}
table(subject$race, subject$biological_sex)
```

## Side-Note: Working with dates
Two of the columns of subject contain dates in the Year-Month-Day format. Recall from our last mini-project that dates and times can be annoying to work with at the best of times. However, in R we have the excellent lubridate package, which can make life allot easier. 

```{r}
library(lubridate)
```

```{r}
today()
```
```{r}
today() - ymd("2000-01-01")
```

```{r}
time_length( today() - ymd("2000-01-01"),  "years")
```

Note that here we are using the ymd() function to tell lubridate the format of our particular date and then the time_length() function to convert days to years.

```{r}
subject$age <- today() - ymd(subject$year_of_birth)
```

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
subject %>%
group_by(infancy_vac) %>%
summarise(mean_age = round(mean(time_length(age, "years")),1))
```
The average age of individuals who received the acellular pertussis vaccine (aP) is approximately 28 years, while the average age of individuals who received the whole-cell pertussis vaccine (wP) is approximately 37 years. The wP group is therefore older on average. This difference is expected because the whole-cell vaccine was used earlier in the United States before being replaced by the acellular vaccine in the 1990s.

> Q8. Determine the age of all individuals at time of boost?

```{r}
subject$age_at_boost <-
ymd(subject$date_of_boost) -
ymd(subject$year_of_birth)
```

```{r}
time_length(subject$age_at_boost, "years")
```

> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
aes(time_length(age, "years"),
fill=as.factor(infancy_vac)) +
geom_histogram(show.legend=FALSE) +
facet_wrap(vars(infancy_vac), nrow=2) +
xlab("Age in years")
```

```{r}
library(httr)
library(jsonlite)

res <- GET("https://www.cmi-pb.org/api/ab_titer/")
titer <- fromJSON(content(res, "text"), simplifyVector = TRUE)
```

```{r}
head(titer)
```


```{r}
specimen <- read_json(
"https://www.cmi-pb.org/api/specimen",
simplifyVector = TRUE)
```

```{r}
titer <- read_json(
"https://www.cmi-pb.org/api/v1/ab_titer",
simplifyVector = TRUE
)
```


## Joining multiple tables
Read the specimen and ab_titer tables into R and store the data as `specimen` and `titer` named data frames.

To know whether a given `specimen_id` comes from an aP or wP individual we need to link (a.k.a. “join” or merge) our `specimen` and `subject` data frames. The excellent dplyr package (that we have used previously) has a family of `join()` functions that can help us with this common task.

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details

```{r}
meta <- inner_join(specimen, subject, by = "subject_id")

dim(meta)
head(meta)
```

```{r}
library(httr)
library(jsonlite)

res <- GET("https://www.cmi-pb.org/api/v1/ab_titer")

titer <- fromJSON(content(res, "text", encoding = "UTF-8"), flatten = TRUE)
```


```{r}
names(titer)
head(titer)
```

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(meta, titer, by = c("specimen_id" = "sample_id"))

dim(abdata)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

> Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

### 4. Examine IgG Ab titer levels
Now using our joined/merged/linked `abdata` dataset `filter()` for IgG `isotype`.

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
aes(ab_titer, antigen) +
geom_boxplot() +
xlim(0,75) +
facet_wrap(vars(visit), nrow=2)
```

> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

The antigen that shows the largest change in IgG antibody titers over time is PRN, which shows a clear increase in antibody levels at later visits compared to earlier visits. PT also shows some increase over time but the change is smaller. FHA shows relatively little change across visits. These differences likely occur because PRN and PT are strong targets of the immune response following pertussis vaccination, leading to increased antibody production after boosting, while FHA does not show as large of a boosted response.