---
title: "Lab_08"
author: "May Alsoufi (PID: A17423250)"
format: pdf
toc: true
---

## Background

In today's class we will be employing all the R techniques for data analysis that we have learned thus far- including the machine learning methods of clustering and PCA- to analyze real breast cancer biopsy data.

## Data import

The data is in CSV format:

```{r}
fna.data <- "WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names=1)
```

Can also be done like this:

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names=1)
```

Wee peak at the data
```{r}
head(wisc.df, 3)
```

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.df)
```

> Q2. How many of the observations have a malignant diagnosis?

```{r}
table(wisc.df$diagnosis)
```
You can also find it specifically like this:

```{r}
sum(wisc.df$diagnosis == "M")
```

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
length(grep("_mean$", colnames(wisc.df)))
```

We need to remove the `diagnosis` column before we do any further analysis of this dataset - we do not want to pass this to PCA etc. We will save it as a separate wee vector that we can use later to compare our finidings to those of experts.

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df$diagnosis
```

## Principal Component Analysis (PCA)

The main function in base R is called `prcomp()` we will use the optional argument `scale=TRUE` here as the data columns/features/dimensions are on very different scales in the original data set.

```{r}
wisc.pr <- prcomp(wisc.data, scale=T)
```

```{r}
attributes(wisc.pr)
```
```{r}
library(ggplot2)
ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```

> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

```{r}
pr.var <- wisc.pr$sdev^2
pve <- pr.var / sum(pr.var)
pve[1]
```

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

```{r}
cumsum(pve)
```

> Q6. How many PCs required to describe at least 90% variance?

```{r}
which(cumsum(pve) >= 0.90)[1]
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

The biplot is cluttered and hard to interpret due to many variables and samples; individual points and vectors overlap heavily — making it difficult to see clear structure.

> Q8. Generate scatter plots of PC1 vs. PC2 and PC1 vs. PC3. What do you notice?

PC1 vs PC2 shows some separation between malignant and benign samples (especially along PC1). PC1 vs PC3 shows much less separation.

## Interpreting PCA results
Now you will use some visualizations to better understand your PCA model. A common visualization for PCA results is the so-called biplot.

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

## Hierarchical Clustering

The goal of this section is to do hierarchical clustering of the **original data** to see if there is any obvious grouping into malignant and benign clusters.

Hierarchical clustering is distance-based, so scaling is essential (same reason as PCA). `dist()` computes pairwise distances between rows (observations) by default.

> Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
wisc.dist <- dist(scale(wisc.data))
wisc.hclust<-hclust(wisc.dist)
plot(wisc.hclust)
```

```{r}
wisc.hclus.clusters <- cutree(wisc.hclust, k=2)
table(wisc.hclus.clusters)
```

> Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

ward.D2 usually gives clearer, more separated clusters because it minimizes within-cluster variance. 

## Combining methods

The idea here is that I can take my new variables i.e. the scores on the PCs `wisc.pr$x`) that are better descriptors of the data-set than the origional features (i.e. the 30 columns in `wisc.data`)

>> Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?

```{r}
pc.dist <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(pc.dist, method="ward.D2")
plot(wisc.pr.hclust)
```

> Q13. How well does this clustering match diagnoses?

From the table:
Group 1: mostly malignant
Group 2: mostly benign

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

I can now run `table()` with both my clustering `grps` and the expert `diagnosis`

```{r}
table(grps, diagnosis)
```

> > Q14. Compare performance of hierarchical clustering models

Our cluster "1" has 179 "malignant" diagnosis. 
Our cluster "2" has 333 "benign" diagnosis.

179 TP.
24 FP.
333 TN.
33 FN.

Sensitivity: Ability to correctly detect ill patients who do have the condition. Equation: TP/(TP+FN)
```{r}
179/(179+33)
```

Specificity: Ability to correctly reject healthy patients without a condition. Equation: TN/(TN+FP)
```{r}
333/(333+24)
```

## Prediction

We will use the `predict()` function that will take our PCA model from before and new cancer cell data and project that data onto our PCA space.

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```
> Q16. Which new patients should be prioritized for follow-up?

```{r}
npc <- predict(wisc.pr, newdata = new)
plot(wisc.pr$x[,1:2], col = grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c("1","2"), col="white")
```
